Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2020-10-28 write-up/covid-writeup.tex
Update covid-writeup.tex

\numberwithin{equation}{section}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{example}{Example}
\newtheorem{remark}{Remark}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}

@@

\usepackage{setspace}
\doublespacing

@@

A molecular test conducted on an infected individual may return a negative result.  Such false negatives are very strongly associated with when the test is conducted.  In the incubation phase, most molecular tests will return a false negative result.  Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility (cite: ).   Moreover, most molecular tests are performed via nasopharyngeal swab (cite: NEJM).  Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.   Systematic reviews suggest that 70\% sensitivity and 95\% specificity are reasonable estimates for current RT-PCR tests (cite: NEJM).

@@

The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.  This raises the immediate question whether prevalence estimators can be improved.  In this section, we consider two directions. First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator. We then extend the statistical error decomposition to the weighted setting. Second, we consider an extension of compartmental models from infectious disease epidemiology using death data.  After adjusting for measurement-error, this leads to an alternative, model-based forecast of prevalence. We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators. We provide an appropriate extension of the statistical error decomposition. The method and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.

@@

To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.  For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now. Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$. For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.  These are discrete factors and therefore define a stratification of the population.  Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.  Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.

Maximum likelihood estimation follows by maximizing
\begin{equation}
\label{eq:propensity}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
\end{equation}
The first sum only involves individuals observed in the nonprobability sample.
The second sum is over the entire population.  Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.  Typically, this is not possible.  Here we present a method that uses auxiliary information obtained from probability samples as well as probability subsamples from the self-selected population.

@@

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
\begin{equation}
\label{eq:auxinfoprob}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
\end{equation}
Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population. Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.  We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the design-unbiased estimator
\begin{equation}
\label{eq:auxinfoprob2}
\sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
\end{equation}
Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.  The estimate now relies on random sampling in both the first and second term.

\subsubsection{An IPW estimator}

Selection propensities can be constructed to use both types of auxiliary information (i.e., combine Sections~\ref{subsec:auxsum} and~\ref{subsec:auxprob}). Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
\begin{equation}
\label{eq:ipwest}
\bar y_n^\star
= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
\end{equation}
where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$. Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.

@@

As   necessary covariates go beyond demographic information, random sampling approaches as discussed in Section~\ref{subsec:auxprob} are required.  If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.  Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.  For simplicity, here we discuss a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
$$
\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
$$
where $K_h$ is a kernel function with tunable parameter $h$. Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
$$
\hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
$$
where $\tilde d_k$ is the number of observations in strata $k$ in the random sample. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.

\subsubsection{Asymptotics and a statistical error decomposition}

Assuming the selection propensity model is correctly specificed, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance:

\begin{lemma}[IPW Variance] \normalfont
\label{lemma:ipw}
Under regularity assumptions specified in the Appendix, we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ (check) where:
$$
V_t^{(IPW)} = \frac{1}{N^2} \sum_{t^\prime = 1}^T \sum_{i=1}^N (1 - \pi_{i,t}) \pi_{i,t} \left( \frac{y_{i,t} - \mu_t}{\pi_{i,t}} -  \right) + b_{1,t}^\top D b_1 + \tilde b_2^\top D_2 b_2.
$$
where
$$
b_{1,t} = \left[ \sum_{t^\prime=1}^T K_h (|t-t^\prime|)\sum_{j=1}^N (1-\pi_{i,t}) (y_{i,t^\prime} - \mu_{t^\prime}) x_{i,t^\prime}) \right] \left[ N^{-1} \sum_{t^\prime=1^T} K_h (|t-t^\prime|)\sum_{j=1}^N (1-\pi_{i,t} (y_{i,t^\prime} - \mu_{t^\prime}) x_{i,t^\prime})\right] $$,
$b_2 = $, $D_1 = N^{-2} V_p ()$, $D_2 = N^{-2} V_p$, and $V_{p,1}$ denotes the design-based variance under the probability sampling designs.

A plug-in estimator is given by
$$
V_t^{(IPW)} = \frac{1}{N^2} \sum_{i=1}^N I_{j,t} (1 - \pi_{i,t}) \left( \right) + \hat b_{1,t}^\top \hat D_t  \hat b_{1,t} + \tilde \hat b_{2,t}^\top \hat D_{2,t}  \hat b_{2,t}.
$$

\end{lemma}

Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
\begin{equation}
\label{eq:statdecomp2}
\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
\end{equation}
where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, and $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution.  See Appendix XX for the derivation.

Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in two ways.  First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.  Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.  Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality.

As demonstrated in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.  Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.  Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation. Interestingly, the impact of measurement-error on data quality is unchanged when considering a weighted estimand.  Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.  In section XX, independent random samples are used to assess effective sample sizes and adjust corresponding CIs.

@@

Selection bias in case count data could be direcly addressed by extending the above compartmental model.  Here, we instead opt to bring in a different source of information by considering COVID-19 death data.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$ and $\nu_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.  Then
$$
D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r p_{k} \nu_{k,t} \theta_{(r-t)} \right)
$$
where $p_{k}$ is the infection fatality rate for the $k$th strata.  This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on age.  Indeed,~\cite{Levin2020} demonstrate that ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85.'' For conciseness, parameter choices and inference are discussed in depth in Appendix XX.

\subsection{Doubly robust estimation.}
Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities, measurement-error and random subsampling of nonprobability samples. In this context, the doubly-robust estimator is given by
$$
\bar y_{n}^{(DR)} = \frac{1}{1 - FP - FN} \left[ \frac{1}{N} \sum_{j=1}^N \hat P_k \hat \mu (x_i) + \frac{1}{\sum_{i=1}^n w(x_i)} \sum_{j=1}^N I_j w(x_i) (Y_i - \hat \mu_k ) \right] - FP.
$$
where $x_i$ is the strata information for individual $i$. This estimator is ``doubly-robust'' because it is robust to either the model-based forecasts or the propensities being correctly specified.  Setting $\pi_i = n/N$ and $\mu(x_i; \hat \beta) = 0$ we again recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.

The key issue with a doubly-robust estimator is the strata information is not observed for every individual in the population. Here, we again use auxiliary information through population-level summary statistics and probability samples and consider the estimator
$$
\frac{1}{1-FP-FN} \left[ \sum_{i=1}^N \hat P_k \hat \mu_k + \sum_{i=1}^n \frac{w_i}{w} \left(y_i - \hat \mu (x_i; \theta) \right) \right] - FP.
$$
where $\hat P_k$ is the estimated fraction in strata $k$.

\subsection{Variance estimator and error decomposition}
A plug-in estimator for the variance
$$
\text{Var} (\bar y_{DR}^\star) =
$$
See Appendix XX for a derivation.

Remark that the robustness is under certain assumptions which we know are likely false; therefore, we are interested in

Trying to decompose this error we start by recognizing $\frac{1}{N} \sum_{i=1}^N Y_i = \sum_k P_k \bar Y_k$.  Then we can concern ourselves with the per strata errors:
$$
\left( P_k - \frac{d_k w_k}{w} \right) \hat y_k + \frac{d_k w_k}{w} \frac{\bar y_k - FP}{1-FP-FN}
$$

{\bf where to describe difference from Chen. (1) weight normalized, (2) use aux info from multiple sources and use raking, (3) use a epi model rather than covariate pred model}.

@@

@@

\begin{subsubsection}{Time-varying doubly-robust prevalence estimation}

@@

\appendix

@@

>>>>

\numberwithin{equation}{section}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
% \newtheorem{example}{Example}
% \newtheorem{remark}{Remark}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}

@@

\usepackage{setspace}
% \onehalfspacing
\doublespacing

@@

A molecular test conducted on an infected individual may return a negative result.  Such false negatives are very strongly associated with when the test is conducted.  In the incubation phase, most molecular tests will return a false negative result.  Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility (cite: that guy ).   Moreover, most molecular tests are performed via nasopharyngeal swab.  Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.   Systematic reviews suggest that 70\% sensitivity and 95\% specificity are reasonable estimates for current RT-PCR tests~\cite{Woloshin2020}.

@@

The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.  This raises the immediate question whether prevalence estimators can be improved.  In this section, we consider two directions.
% First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator. We then extend the statistical error decomposition to the weighted setting. Second, we consider an extension of compartmental models from infectious disease epidemiology using death data.  After adjusting for measurement-error, this leads to an alternative, model-based forecast of prevalence. We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators. We provide an appropriate extension of the statistical error decomposition.
The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.

@@

To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.  For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now. Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$. Maximum likelihood estimation follows by maximizing
\begin{equation}
\label{eq:propensity}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
\end{equation}
The first sum only involves individuals observed in the nonprobability sample.
The second sum is over the entire population.  Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.  Typically, this is not possible.  Here we present a method that uses auxiliary information obtained from probability samples as well as probability subsamples from the self-selected population.

@@

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
\begin{equation}
\label{eq:auxinfoprob}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
\end{equation}
Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population.
% Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

% Equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.  We can extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding the pseudo-likelihood
% \begin{equation}
% \label{eq:auxinfoprob2}
% \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
% \end{equation}
% Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.  Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.

\subsubsection{An IPW estimator and statistical error decomposition}
\label{section:IPWerrordecomp}

Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
\begin{equation}
\label{eq:ipwest}
\bar y_n^\star
= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
\end{equation}
where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$. Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.

Let $\tilde I_j (X_j) = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
\begin{equation}
\label{eq:statdecomp2}
\rho_{\tilde I (X), Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{\tilde D_M}
\end{equation}
where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $I_J = 1$, $\rho_{\tilde I(X), Y}$ is the empirical correlation which here depends on covariate distribution, $\tilde f_k = E[ W_J I_J \mid Y_j = k]$ for $k=0,1$, and $\tilde \Delta = \tilde f_1 - \tilde f_0$.  See Appendix~\ref{app:ipwderivation} for the derivation.

Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.  First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.  Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.  Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality. Third, the impact of measurement-error on data quality is changed when considering a weighted estimand. {\red state that it may switch sign!als}

As demonstrated in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.  Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.  Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.   Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.

@@

If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.  Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.  For simplicity, here we discuss a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
$$
\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
$$
where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.

% Under simple random sampling of equal size~$\tilde n$ from a fixed population size~$N$ at each time and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the
% $$
% \hat \pi_{k,t} = \frac{n}{N} \times \frac{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \hat d_{k,t^\prime}}{\sum_{t^\prime=1}^T K_h(|t^\prime - t|) \tilde d_{k,t^\prime}}
% $$
% where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

\subsubsection{Asymptotics}

Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:ipw} shows that the IPW estimator is  consistent and calculates the estimator's variance.  In Lemma~\ref{lemma:ipw}, sensitivity and specificity are not assumed known.  Instead, we suppose that these are estimated using a pseudo-likelihood on a random sample from the population whose size is chosen so the resulting estimated sensitivity and specificity have desired levels of uncertainty.  See Appendix~\ref{app:asympderivations} for additional details.

\begin{lemma}[IPW Variance] \normalfont
\label{lemma:ipw}
The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

@@

Selection bias in case count data could be direcly addressed by extending the above compartmental model.  Here, we instead opt to bring in a different source of information by considering COVID-19 death data.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$ and $\nu_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.  Then
$$
D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r p_{k} \nu_{k,t} \theta_{(r-t)} \right)
$$
where $p_{k}$ is the infection fatality rate for the $k$th strata.  This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on age.  Indeed,~\cite{Levin2020} demonstrate that ``the estimated IFR is close to zero for children and younger adults but rises exponentially with age, reaching 0.4\% at age 55, 1.4\% at age 65, 4.6\% at age 75, and 15\% at age 85.'' For conciseness, parameter choices and inference are discussed in depth in Appendix~\ref{app:modelbased}.

\subsection{Doubly robust estimation.}
Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error. In this context, the doubly-robust estimator is given by
$$
\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
$$
where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data. This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified. A statistical error decomposition can be derived
\begin{align}
\label{eq:statdecomp3}
\begin{split}
\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
\end{split}
\end{align}
See Appendix~\ref{app:DRderivation} for the derivation. Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based focast impacts the estimation error in two ways.  First, if the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components. Note that the model-based forecast does not impact the data quantity component.

If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size. Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.  For decent models, one would expect this ratio to be greater than one.  Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.  Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.

Unfortunately, the first term cannot be computed as covariate information is not collected on the entire population.  Here, we again use the probability samples and consider the doubly-robust estimator
$$
\frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
$$
where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit. Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the IPW estimator is  consistent and calculates the estimator's variance. See Appendix~\ref{app:asympderivations} for additional details.

\begin{lemma}[DR Variance] \normalfont
\label{lemma:dr}
The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

@@

% For COVID-19 case-count data, the only covariate information provided by state and local governments are summary statistics for those factors discussed in Section~\ref{section:testinginfo} -- age, sex,  and race.  These are discrete factors and therefore define a stratification of the population.  Currently, US testing does not collect additional covariate information such as symptom information; however, this need not be the case moving forward.  Recommendation XX below states that additional covariate information should be collected from nonprobabilistic samples.

@@

\subsubsection{Time-varying doubly-robust prevalence estimation}

@@

\newpage
\appendix

@@

\section{IPW statistical error decomposition derivation}
\label{app:ipwderivation}
We start by considering the empirical weighted mean estimator under imperfect testing,
$$
